Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-04-12 analysis/paper/paper.Rmd
In progress: writing

---
title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

@@

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, these models generally do not include road network projections. For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) project forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then add fire. This example highlights that roads are also important for harvest planning, and that road network projection methods are integrated into a widely used optimization-based spatial harvest planning tool. However, Patchworks is proprietary closed-source commercial software, and the road network projection algorithm contained therein is difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Published methods intended to support strategic planning [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] require inputs that are not publicly available, and employ algorithms that are too complex and detailed for application at regional, provincial or national scales. More promising options for projecting forest road network development at large scales include an iterative least cost path method developed by Hardy [-@hardy_forest_2019], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis plugins, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. These methods have not been formally assessed, compared, published, or implemented in a way that allows for easy integration into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016; @scheller_forest_2004].

In this work, we address the need for adequately accurate forest road network projection tools that are also widely and easily applicable by implementing simple iterative least cost path and minimum spanning tree methods in an open source R package. We compare the ability of these methods to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia using performance metrics relevant to wildlife. The adequacy of a road network projection method depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose. Common predictors in models of the impacts of roads on wildlife include distance to roads [], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011], and density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017]. We expect projecting these measures to be less challenging than projecting the spatial location of each road segment, and we do not expect good projections of road locations from the algorithms considered here. Our goal is to assess the ability of simple iterative least cost path and minimum spanning tree methods to project footprint, distance and density metrics that are commonly used to predict impacts of roads on wildlife.

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) (Figure \@ref(fig:proj-meth-fig)). We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock are determined independently in the simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic, such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connect all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:proj-meth-fig)). We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019].

@@

To help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package `roads` [@endicott2023]. The 'roads' package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi2006a; @igraph:2022], `terra` for manipulating spatial raster data [@hijmans2023], and `sf` for manipulating spatial vector data [@pebesma2018; @pebesma2022]. The package is available on CRAN (The Comprehensive R Archive Network) and a website with tutorials and detailed documentation is available (<https://landscitech.github.io/roads/>). The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>).

@@

## Aspatial performance
For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). All road network projection method variants represent an improvement over the "cutblocks only" method where cutblocks were the only disturbance considered. Within all cutblocks, the Hardy method most accurately projects road density, and least accurately projects distance to road. Low density method variants most severely underproject road density within cutblocks, and most accurately project distance to road within cutblocks. These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g. Figure \@ref(fig:proj-map-fig)b). Omitting cutblocks without observed roads from the comparison (Figure \@ref(fig:aspat-perf-fig) row 2) clarifies that within cutblocks the Hardy method more accurately projects ... All methods underproject road density, in part because within 1 ha pixels all the methods generate straight road segments. All the methods also underproject road presence, and the problem is more severe for the low density method variants. In general, performance differences among low and high density method variants are larger than differences among projection algorithms. With the coarser resolution cost raster road projections had mean road presence that was closer to the observed, while road density was further from the observed mean and distance to the nearest road was further from the observed mean inside cutblocks but closer for the landscape overall (Figure A-\@ref(fig:figA1)).

Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig)). Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a). Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms could likely reduce this discrepancy. Another source of discrepancy between observed and projected networks in this mountainous landscape are switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c). None of the simple algorithms considered here are capable of projecting switchbacks. Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project.

@@

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy highlighted in this mountainous terrain include switchbacks, stream crossings, inaccurate harvest locations. we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield shows better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. In particular, none of these simple algorithms will project switchbacks on steep slopes. A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. 
- Aspatial overall performance metrics can be difficult to interpret. To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors. Better harvest & existing road data would allow us to distinguish errors in projections from errors in the input data (cite lit in coops paper)
 - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing. Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm, and the Hardy algorithm more accurately projects road density than either. Hardy algorithm is also less prone to putting roads on block edges. If accurate road density within blocks is primary goal, consider using Hardy algorithm. Thus far only implemented in Landis II and QGIS extension, but source code is open. If there is a need could integrate into roads R package. Or consider simpler adjustments to algorithms for selecting landing points such as placing landing points away from cutblock edges. 
- However, before investing in more complexity recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates (). In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). If our wildlife response models depend on distance to road or buffered disturbance footprint, these somewhat crude road projections are likely good enough.
- A related point is that it may helpful to recognize that some road metrics are easier to project than others when constructing model to project wildlife responses to disturbance ... (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.

>>>>

---
title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
    equal_contributor: "yes"
    orcid_id: 0000-0001-7875-9015
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
    equal_contributor: "yes"
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

@@

Increasingly rapid environmental change, improvements in available cyberinfrastructure, and development of open source data, tools and methods have contributed to growing interest in ecological forecasting models [@bodner_bridging_2021; @dietze_ecological_2017; @mcintire_perfict_2022]. In particular, forest landscape models that can account for harvest, fire, and other disturbances are widely and increasingly used to project the cumulative effects of climate change and forest management [@petter_how_2020; @albrich_simulating_2020; @hof_editorial_2021; @tremblay_harvesting_2018; @snell_importance_2018; @daniel_incorporating_2017; @stewart_climate-informed_2023; @micheletti_assessing_2021; @leblond_there_2022; @bouderbala_long-term_2022; @barros_empowering_2023; @labadie_global_2023], building on a rich history of spatial-stochastic models of forest growth and succession following disturbance [@sturtevant_understanding_2021]. Despite the importance of roads for wildlife, these models generally do not include road network projections. For example, in a recent collection of articles focussed on "Using Landscape Simulation Models to Help Balance Conflicting Goals in Changing Forests" [@hof_editorial_2021], only 4 of 14 mentioned roads [@haga_scenario_2020; @leston_quantifying_2020; @mikusinski_strengthening_2021; @norris_forecasting_2021], and only one includes a road projection [@leston_quantifying_2020].

Leston et al. (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire. This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] and tools exist [@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate; @epstein_combinatorial_2006]. However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Methods intended to support tactical and operational planning also require detailed parameterization with inputs that may not be publicly available, and employ algorithms that are too complex for application at regional, provincial or national scales. More promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. Hardy's tool is designed for integration with other components of the Landis-II modelling framework [e.g. @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

To help address the need for adequately accurate forest road network projection tools that can be easily integrated into stochastic models of forest change we implemented simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in an open source R package. We compared the ability of these methods and Hardy's method [-@hardy_forest_2019; -@hardy_landis-ii_2023] to reconstruct observed forest road network development since 1990 in a mountainous region of British Columbia. The adequacy of a road network projection methods depends on the purpose of the projection, and it is important to select performance metrics that reflect the purpose. Our interest is projecting impacts of forest change on wildlife, so we focused on common predictors of the impacts of roads on wildlife, including distance to roads [@labadie_global_2023], buffered disturbance footprint [@johnson_science_2020; @ibisch_global_2016; @lochhead_linking_2022; @polfus_identifying_2011; @labadie_global_2023], and road density [@apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017].

@@

Simple methods for projecting low-cost routes from an existing road network to target locations (i.e. harvest cutblocks) include simple least cost paths (LCP), dynamic least cost paths (DLCP) and least cost path minimum spanning trees (MST) (Figure \@ref(fig:proj-meth-fig)). We use Dijkstra's shortest path algorithm [-@dijkstra_note_1959] to determine the least cost path between a target location (e.g. centroid of a cutblock) and the existing road network. Least cost paths to each cutblock are determined independently in the simple method (LCP). The dynamic least cost paths (DLCP) method evaluates the path to each cutblock in sequence so that earlier paths are considered in the cost for later paths. Sequence is determined by a heuristic, such as the distance to the nearest existing road in these examples. The minimum spanning tree (MST) method calculates the minimum cost network to connect all cutblocks to the road network using Kruskal's algorithm [-@kruskal_shortest_1956]. See Appendix A for a more detailed explanation of these methods. We focus our analysis on the DLCP and MST approaches which yield more realistic road networks (Figure \@ref(fig:proj-meth-fig)). We compared these methods to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023].

@@

To help make these road network projection methods more easily and widely applicable, we implemented the simple least cost path, dynamic least cost path and minimum spanning tree methods in the open source R package `roads` [@endicott_roads_2023]. The 'roads' package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network) and a website with tutorials and detailed documentation is available (<https://landscitech.github.io/roads/>). The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>).

@@

## Aspatial performance
For each projection method variant the aspatial performance varied across the five road network metrics and also varied depending on whether we consider scores within cutblocks, outside cutblocks, or overall (Figure \@ref(fig:aspat-perf-fig)). Within all cutblocks, the Hardy method most accurately projected road density, and least accurately projected distance to road. Low density method variants most severely underprojected road density within cutblocks, and most accurately projected distance to road within cutblocks. These apparently contradictory results are explained by areas in our input data that have been misidentified as cutblocks with no associated observed roads (e.g. Figure \@ref(fig:proj-map-fig)b). Omitting cutblocks without observed roads from the comparison (Figure \@ref(fig:aspat-perf-fig) row 2) clarifies that within cutblocks the Hardy method most accurately projected road density and distance to road. However, note that our observed landscape is densely roaded in part because switchbacks are required in steep terrain (Figure \@ref(fig:proj-meth-fig)). The Hardy method projects more wiggles than the other methods, but none of the methods considered here can project switchbacks (Figure \@ref(fig:proj-meth-fig)).

Another fundamental limitation of all these methods is that they generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the cost raster resolution is coarse coarse (Figure A-\@ref(fig:figA1)). In contrast, projections of the proportion of cells that contain roads is more accurate on a coarse raster (Figure A-\@ref(fig:figA1)).

Outside cutblocks, all method variants underproject road density and road presence, and there is little variation in performance among them (Figure \@ref(fig:aspat-perf-fig)). Comparison of observed and projected road networks clarifies that projected roads cross valley bottoms far more often than observed roads which often run parallel to one another on each side of rivers (Figure \@ref(fig:proj-map-fig)a). Accounting for the costs of crossing streams, rivers and other wet areas in valley bottoms could likely reduce this discrepancy. Switchbacks on steep slopes (Figure \@ref(fig:proj-map-fig)c) are also a source of discrepancy outside of cutblocks. Overall, projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. Performance differences among low and high density method variants are larger than differences among projection algorithms, and all road network projection method variants represent an improvement over the "cutblocks only" method where cutblocks were the only disturbance considered.

@@

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models). Our R package provides a comparable alternative to Hardy QGIS and Landis plugins, more easily integrated into SpaDES or other R workflows. 
- as expected, none of these algorithms are great at projecting exact road locations. we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield shows better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. In particular, none of these simple algorithms will project switchbacks on steep slopes. A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. 
- Aspatial overall performance metrics can be difficult to interpret. To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors. Better harvest & existing road data would allow us to distinguish errors in projections from errors in the input data (cite lit in coops paper)
 - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing. Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm. Hardy algorithm most accurately projected road density within blocks - could up density of landing pts in DCLP/MST algorithms to match this. But note that these algorithms can get the right density without replicating the pattern or mechanism. Hardy density is high in part because the algorithm is prefering turns over straigt lines straight lines, which may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. Projected wiggles are quite different from observed switchbacks. 
- If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do. But before investing in more complexity recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates (). In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). If our wildlife response models depend on distance to road or buffered disturbance footprint, these somewhat crude road projections are likely good enough.
- A related point is that it may helpful to recognize that some road metrics are easier to project than others when constructing model to project wildlife responses to disturbance ... (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
